---
title: 'MA40198: Applied Statistical Inference'
author: 
- Candidate numbers 07604, 07614, 07615

date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1: Carcinogenesis Study on Rats

### Abstract
* Brief outline of: the problem; what you have done; and your conclusions. 
* Try not to use symbols/numbers 
* concise
* this needs editing  

Following a drug trial that was carried out on 150 rats, this study aims to find out if the cancer treatment which has been used on is effective or not.  

### Introduction

* This should describe the general context and background, including a general description of what data are available, the manner in which they were collected. 
* Talk about aims of the investigation 
* Indicate the methods used
* Expanded version of the abstract
* Maybe we could include separations of the data i.e. look at specifically how many had tumours etc etc
* Needs editing pls

The recorded data includes the variables; *Litter*, observing 3 rats from each of the 50 different litters;  *Treatment Indicator*, where rx=1 indicates treatment was carried out on the rat and rx=0 indicates the control; *Time*, the follow up time in weeks to the appearance of a tumor; *Status*, which describes whether there was presence of a tumour status=1 or not status=0. 

This study is comprised of two main approaches of statistical analyses: frequentist estimation procedures and Bayesian estimation procedures. Using both the Bayesian and Frequentist approach, probability models for the follow up time and the survival function have been distributed using the Weibull function. A Frequentist approach was also taken whilst modelling the data using a log-logistic distribution. 

### Methods

* Use fairly detailed explanation of methods incl any theory necessary
* The reader should be able to repeat the study on the basis of the report.
* Could split the methods section into 
+ a pleliminary analysis, using graphical methods/ looking at simple descriptive methods if needed???
+ then a full scale analysis
* and then second section (Maybe into two subsections)
+ a description, explanation or development of the methodology used (this should be the most technical part of the report. It should describe detailed precise mathematical tools & techniques used)
+ the actual analysis.
* Include the graphs to show how we obtained the starting points maybe as 3 in a row. i.e. par(mfrow=c(1,2))
* The code for the functions

### Results, conclusions and recommendations
* Give the main results and conclusions. 
* Try to report your conclusion in context of the experiment from which the data came.

### References
?? Maybe not

### Appendices
Maybe not??????

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
